Neural Connectome of the Ctenophore Statocyst

### 2025-04-25 manuscript/Manuscript.qmd
edit fig1

In cells containing cilia, we also traced cilia along their length and annotated all basal bodies. For neuronal skeletonization, nodes were placed to interconnect the profiles of the same neuron’s processes across layers, extending the skeleton until all branches were fully traced. Each node was tagged <!--# is this true? not only a subset?  -->, and skeletons were named and assigned multi-level annotations. As described later, some neurons formed loop-like structures (anastomosed neurons), wherein separated branches often rejoined either the main branch or smaller branches [@Burkhardt2023]. In such cases, branch nodes were placed near the closest existing node and annotated accordingly (CATMAID only supports skeleton trees). The entire skeletonized volume was composed of 134,591 nodes. 88 fragments could not be attached to somata-associated skeletons. Most of these fragments represent short skeletal branches that could not be traced beyond gaps or low-quality layers.

@@

![**Figure 1. Morphology of the aboral organ in *Mnemiopsis leidyi***\
(A) Whole-body image of a 5-day-old *M. leidyi* cydippid larva in lateral view. The boxed region indicates the aboral organ (ao). Scale bar: 100 µm.\
(B) Schematic diagram of the aboral view of a cydippid larva. The two-tone coloration represents the biradial symmetry of the body. For convenience, the body is divided into four quadrants, designated as the first (Q1) through fourth (Q4) quadrants. The two primary viewing angles are referred to as the sagittal plane (S) and the tentacular plane (T). Abbreviations: ao, aboral organ; cg, ciliated groove; cr, comb plates; tb, tentacle bulbs.\
(C) Enlarged views of the aboral organ from different perspectives in a cydippid larva using a DIC microscope. The left panel presents an aboral view of the aboral organ. The middle panel shows a lateral view of the aboral organ in the sagittal (S) plane, while the right panel displays a lateral view in the tentacular (T) plane. Abbreviations: bal, balancer; cg, ciliated grooves; do, dome cilia; li, lithocyte. Scale bar: 50 µm.\
(D) Schematic representations of the aboral organ from three different perspectives. The left panel illustrates the aboral view, the middle panel shows the lateral view in the sagittal plane, and the right panel presents the lateral view in the tentacular plane. The quadrant colors correspond to those in panel (B).\
(E) Example of cell tracing using the collaborative annotation toolkit CATMAID for large-scale electron microscopy image datasets. The spherical objects indicate nuclear positions, while the lines represent the traced cell centers. Scale bar: 10 µm.\
(F) Three-dimensional reconstruction of cells composing the aboral organ, displayed from different perspectives. The left panel presents an aboral view of the aboral organ. The middle panel shows a sagittal plane view, while the right panel provides a tentacular plane view. Cells are color-coded according to their respective quadrants. Lithocytes (li) are represented as three gray spheres, while balancers (bal) are depicted as gray lines. Scale bar: 25 µm.](figures/Figure1.png)

@@

Larvae of *Mnemiopsis leidyi* (five days old) were cryofixed using a high-pressure freezing apparatus (BAL-TEC HPM 010, Balzers) <!-- was this done in Exeter or Bergen? If Exeter, it is not the Baltec --> and immediately transferred to liquid nitrogen for storage. The frozen samples were processed in a substitution medium containing 2% (w/v) osmium tetroxide and 0.5% uranyl acetate in acetone, using a cryo-substitution device (EM AFS-2, Leica). Cryo-substitution was carried out by gradually raising the temperature  <!-- steps? solutions? --> , and the samples were embedded in epoxy resin <!--type? -->. Serial untrithin sections of 50 nm thickness were prepared using a Reichert Jung Ultracut E ultramicrotome and a 45º DiATOME diamond knife <!-- wasn't it our Leica UC7? -->. To enhance section adhesion and improve hydrophilicity, a conductive indium tin oxide-coated glass slide (ITO Glass, UQG Optics) was treated with air glow discharge using the PELCO easiGlow system (Ted Pella, Inc.), rendering the carbon film <!-- was there carbon film? --> surface negatively charged. Section ribbons were collected on the prepared glass slides, slowly dried to ensure proper stretching, and firmly adhered to the glass surface. The sections on glass were stained with UranyLess and lead citrate (Reynolds) using airless staining bottles (Delta Microscopies). The glass slides were mounted on STEM-specific stages (Zeiss) using Copper Foil EMI Shielding Tape (3M). Imaging was done on a Gemini SEM 500 (Zeiss) equipped with SmartSEM and Atlas 5 imaging software (Zeiss). The full dataset consisted of 620 serial sections, each 50 nm thick. The imaging resolution was 2.8 nm/pixel, using electron holography transmission <!-- is this true? --> at an acceleration voltage of 1.5 kV (in-lens detector, dwell time: 3 µs  <!-- sounds very low -->).

>>>>

In cells containing cilia, we also traced cilia along their length and annotated all basal bodies. For neuronal skeletonization, nodes were placed to interconnect the profiles of the same neuron’s processes across layers, extending the skeleton until all branches were fully traced. All nodes relevant to synaptic sites were tagged, and skeletons were named and assigned multi-level annotations. As described later, some neurons formed loop-like structures (anastomosed neurons), wherein separated branches often rejoined either the main branch or smaller branches [@Burkhardt2023]. In such cases, branch nodes were placed near the closest existing node and annotated accordingly (CATMAID only supports skeleton trees). The entire skeletonized volume was composed of 134,591 nodes. 88 fragments could not be attached to somata-associated skeletons. Most of these fragments represent short skeletal branches that could not be traced beyond gaps or low-quality layers.

@@

![**Figure 1. Morphology of the aboral organ in *Mnemiopsis leidyi***\
(A) Whole-body image of a 5-day-old *M. leidyi* cydippid larva in lateral view. The boxed region indicates the aboral organ (ao). Scale bar: 100 µm.\
(B) Schematic diagram of the aboral view of a cydippid larva. The two-tone coloration represents the biradial symmetry of the body. For convenience, the body is divided into four quadrants, designated as the first (Q1) through fourth (Q4) quadrants. The two primary viewing angles are referred to as the sagittal plane (S) and the tentacular plane (T). Abbreviations: ao, aboral organ; cg, ciliated groove; cr, comb plates; tb, tentacle bulbs.\
(C) Aboral views of the aboral organ highlighting its spatial organization. The left panel presents a schematic representation of the aboral organ, illustrating the four body quadrants (Q1–Q4), color-coded consistently with panel (B). The right panel shows a corresponding differential interference contrast (DIC) image from the same perspective. The region enclosed by the dotted line delineates the boundary of the aboral organ. The sagittal (S) and tentacular (T) arrows indicate the viewing directions corresponding to those planes. Scale bar: 50 µm. bal, balancer; cg, ciliated grooves; li, lithocyte.\
(D) Lateral views of the aboral organ in two orthogonal planes captured using DIC microscopy. The left panel shows a lateral view in the sagittal (S) plane, while the right panel displays a view in the tentacular (T) plane. Balancer cilia (bal) and lithocytes (li) are enclosed within the dome (do). Scale bar: 50 µm. A–O indicates the aboral–oral body axis.\
(E) Schematic representations of the aboral organ in sagittal and tentacular planes.
The left panel shows a lateral schematic in the sagittal plane (S), and the right panel shows the same in the tentacular plane (T), corresponding to the views in panel (D). Colors match the quadrant scheme shown in panels (B) and (C).\
(F) Example of cell tracing using the collaborative annotation toolkit CATMAID for large-scale electron microscopy image datasets. The spherical objects indicate nuclear positions, while the lines represent the traced cell centers. Scale bar: 10 µm.\
(G) Three-dimensional reconstruction of cells composing the aboral organ, displayed from different perspectives. The left panel presents an aboral view of the aboral organ. The middle panel shows a sagittal plane view, while the right panel provides a tentacular plane view. Cells are color-coded according to their respective quadrants. Lithocytes (li) are represented as three gray spheres, while balancers (bal) are depicted as gray lines. Scale bar: 25 µm.](figures/Figure1.png)

@@

Larvae of *Mnemiopsis leidyi* (five days old) were cryofixed using a high-pressure freezing apparatus (HPM Live μ, CryoCapCell, https://www.cryocapcell.com/technologies/high-pressure-freezer) and immediately transferred to liquid nitrogen for storage. The frozen samples were processed in a substitution medium containing 2% (w/v) osmium tetroxide and 0.5% uranyl acetate in acetone, using a cryo-substitution device (EM AFS-2, Leica). Cryo-substitution was carried out by gradually raising the temperature  <!-- steps? solutions? --> , and the samples were embedded in epoxy resin <!--type? -->. Serial untrithin sections of 50 nm thickness were prepared using a Reichert Jung Ultracut E ultramicrotome and a 45º DiATOME diamond knife <!-- wasn't it our Leica UC7? -->. To enhance section adhesion and improve hydrophilicity, a conductive indium tin oxide-coated glass slide (ITO Glass, UQG Optics) was treated with air glow discharge using the PELCO easiGlow system (Ted Pella, Inc.), rendering the carbon film <!-- was there carbon film? --> surface negatively charged. Section ribbons were collected on the prepared glass slides, slowly dried to ensure proper stretching, and firmly adhered to the glass surface. The sections on glass were stained with UranyLess and lead citrate (Reynolds) using airless staining bottles (Delta Microscopies). The glass slides were mounted on STEM-specific stages (Zeiss) using Copper Foil EMI Shielding Tape (3M). Imaging was done on a Gemini SEM 500 (Zeiss) equipped with SmartSEM and Atlas 5 imaging software (Zeiss). The full dataset consisted of 620 serial sections, each 50 nm thick. The imaging resolution was 2.8 nm/pixel, using electron holography transmission <!-- is this true? --> at an acceleration voltage of 1.5 kV (in-lens detector, dwell time: 3 µs  <!-- sounds very low -->).
